Online verification of Commutativity

### 2020-05-19 paper/paper.tex
went over paper

\begin{document}
%
\title{Online verification of Commutativity}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{First Author\inst{1}\orcidID{0000-1111-2222-3333} \and
Second Author\inst{1}\orcidID{1111-2222-3333-4444} \and
Third Author\inst{1}\orcidID{2222--3333-4444-5555}}
%
\authorrunning{F. Author et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Cornell University, Ithaca NY 14853, USA
\email{\{abc,lncs\}@cornell.edu}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
Commutative diagrams arise in many systems, such as an implicit type conversion graph in some programming languages.
It can be useful in program analysis to verify that a diagram representing conversions remains commutative.
A naive approach to solving the problem would take factorial time, and previous work provides an algorithm for the special case of acyclic diagrams that runs in $O(|V|^4|E|^2)$ time, where V is the set of vertices in the diagram, and E, the set of edges. 
We present an algorithm that runs in $O(|V|^2|E| + |V|^3)$ time. We evaluate the algorithm gainst baselines, and see its practical application in the compiler of a domain specific language for geometry types.

\end{abstract}
%
%
%
Many systems can be understood as a category, with a collection of sets, and some transformation functions between the sets- for example, in programming languages, type systems with coercion. If a directed graph is constructed with the sets as nodes and functions as edges between the nodes, then by construction, in the new graph, choice of path should not matter- any choice of path should yield the same transformation between any two nodes. The idea is that each object has a unique representation in a given set in the system, so the transformation mapping between any two nodes should not depend on choice of path used to obtain the transformation. We will use the term "diagram" to refer to such graphs with their associated sets (vertices) and transformations (edges).

Since such commutative diagrams arise in many systems, it is often useful to be able to perform verification and ensure the graphs in question are indeed path independent. 
We formalize the problem as checking the commutativity of a diagram incrementally.
A diagram is said to commute when on selecting any two vertices, any directed path leads to the same result by composition.
Naively checking if all paths in a given graph are path independent could require a number of function equality checks as bad as factorial in the number of nodes, since a path consists of an ordering of nodes. Previous work has identified an $O(|E|^2|V|^4)$ algorithm to verify that an acyclic diagram commutes.

We present a polynomial time $O(|E||V|^2+|V|^3)$ algorithm to verify that commutativity of a graph is maintained over the course of online addition, assuming an oracle to check the abstract notion of equality of functions. The number of calls to the equality checking oracle required by the algorithm is asymptotically tight. With an additional $O(|V|^4)$ optimization step, is is possible to use instead the minimal possible number of equality checks, should these checks be expensive. We compare our solution to three baselines: a naive checking of all path pairs, with only special handling for cycles, a check for all path pairs that involve the new edge, and with an algorithm suggested by previous work to solve the batch version of the problem for acyclic diagrams.

@@

First, we formalize the notion of a diagram, drawing inspiration from Murota\cite{commutative}.

We start with a directed graph G=(V,E) with sets attached to vertices and maps to arcs. Each edge (u, v) in E corresponds to a function that maps elements of u to elements in v.
All the functions form a semigroup F, with multiplication denoted by $\circ$.
A semigroup is an algebraic structure with a set and an associative binary operation. We use it to represent functions in order to capture the one allowed operation- function composition.

@@

Two situation can be imagined- the first, where the existence of the identity function from every node to itself is assumed, and the second, where no such identity needs to exist.

@@

\paragraph{Implementation}
*Copy in python code from implementation*

@@

\subsection{Optimal solution for batch problem}
Murota's main result \cite{commutative} is a solution to the question of whether a given acyclic diagram commutes that returns the minimal number of equality checks. The paper describes the following algorithm to find the ($V^2E$ bounded) minimal set of pairs that needs to be checked.

@@

A bilinking is a parallel pair that is disjoint but for their terminal nodes. The set of all bilinkings is $R_0$.
In an acyclic diagram, if all bilinkings are equal, all parallel pairs must also be equal since they any given pair can be expressed as a composition of one or more bilinkings.

@@

for a set of bilinkings r, the cl function is defined as:
cl(r) = $\{ s\in R_0|$ s is linearly dependent on $F(r) \}$.

The closure function on r basically captures all the pairs that can be made by made by composing or "gluing together" the bilinkings in r. The $\langle\rangle$ function captures the orientation of paths, and can be used to constrain compositions to be in the right direction.

Using these two function we finally define the function $\sigma$ on a set of bilinkings r as
$\sigma(r) = \{s \in R_0 | s\in cl(R\cap \langle s \rangle) \}$
This is the function used to capture all the pairs whose verification is implied by the verification of pairs in r.
It is used to iteratively check if a given pair is redundant. bilinkings are eliminated until a minimum "spanning" subset is reached.

@@

The solution presented here assumes that there is an implicit identity function on every node, as would be the case in most applications. However, a modified version of the algorithm for diagrams without this assumption is presented in the appendix, with modification that is similar in spirit to what was described in the description of the naive baselines.

@@

The try block is executed at most $O(|V|^2)$ times.

@@

The overall run time of the algorithm, assuming that path finding is done using depth first search, is $O((|V|+|E|)|V|^2)$.

@@

The run time of the first step is $O(|V|^4)$, and that of the second step is $O(|V|)$, so that the overall bound is $O(|V|^4)$, which is still a gain of a factor of $|E|^2$ over the batch algorithm.

@@

Set parallelPairs = new Set();
for S in existingGraph.Nodes():
    for T in existingGraph.Nodes():
        try:
            Path pathWithNewEdge = 
                FindPath(
                    graph: existingGraph, 
                    sourceNode: S,
                    sinkNode: newEdge.Source()) +
                newEdge as Path +
                FindPath(
                    graph: existingGraph, 
                    sourceNode: newEdge.Sink(), 
                    sinkNode: T)
            if (S==T):
                // Cycles are a special case.
                parallelPairs.add(pathWithNewEdge, pathWithNewEdge+pathWithNewEdge)
                for edge in (S.forwardEdges - pathWithNewEdge[0]):
                    parallelPairs.add(edge, edge+pathWithNewEdge)
                for edge in (S.backwardEdges - pathWithNewEdge[-1]):
                    parallelPairs.add(edge, pathWithNewEdge+edge)
            else:
                Path pathInOldGraph = FindPath(
                    graph: existingGraph, 
                    sourceNode: S, 
                    sinkNode: T);
                parallelPairs.add((pathInOldGraph, pathWithNewEdge));
        catch (pathFindingFailed Exception):
            // No comparable pairs from node {S} to node {T} 
            // that need to be checked
            continue
    for cycle in findCyclesWithTerminatNode(newEdge.Source()):
        parallelPairs.add(newEdge, cycle+newEdge)
    for cycle in findCyclesWithTerminatNode(newEdge.Source()):
        parallelPairs.add(newEdge, newEdge+cycle)
output parallelPairs;
\end{verbatim}

The optional minimization algorithm is modified to exclude nodes representing cycles, since the reduction case no longer will apply in the absence of an implicit identity on the terminal node of the cycle.
*I think the minimality of minimized subset still holds, but not 100\% sure*

>>>>

\begin{document}
%
\title{Online verification of Commutativity}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{First Author\inst{1}\orcidID{0000-1111-2222-3333} \and
Second Author\inst{1}\orcidID{1111-2222-3333-4444} \and
Third Author\inst{1}\orcidID{2222--3333-4444-5555}}
%
\authorrunning{F. Author et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Cornell University, Ithaca NY 14853, USA
\email{\{abc,lncs\}@cornell.edu}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
Commutative diagrams arise in many systems, such as an implicit type conversion graph in some programming languages.
It can be useful in program analysis to verify that a diagram representing conversions remains commutative over the course of its construction.
A naive approach to solving the problem would take factorial time, and the presence of cycles could potentially lead to an infinite number of paths to check. Previous work provides an algorithm for the special case of acyclic diagrams that verifies that a given diagram commutes in $O(|V|^4|E|^2)$ time, where V is the set of vertices in the diagram, and E, the set of edges. 
We present an algorithm that given a diagram and an edge to add to it, verifies that the resultant diagram commutes in $O(|V|(|E| + |V|)$ time. For the case when checking the equality of paths is expensive, we present a second algorithm that runs in $O(V^4)$ time but uses the minimum possible number of equality checks. We evaluate the algorithms against baselines, and see its practical application in the compiler of a domain specific language for geometry types. As a demonstration of the more general applicability of the algorithm, we use it to identify discrepancies in a  currency conversion graph.

\end{abstract}
%
%
%
Many systems can be understood as a category, with a collection of sets, and some transformation functions between the sets- for example, in programming languages, type systems with coercion. If a directed graph is constructed with the sets as nodes and functions as edges between the nodes, then by construction, in the new graph, choice of path should not matter- any choice of path should yield the same transformation between any two nodes. Each object has a unique representation in a given set in the system, so the transformation mapping between any two nodes should not depend on choice of path used to obtain the transformation. We will use the term "diagram" to refer to such graphs with their associated sets (vertices) and transformations (edges).

Since such commutative diagrams arise in many systems, it is often useful to be able to perform verification and ensure the graphs in question are indeed path independent. 
We formalize the problem as checking the commutativity of a diagram incrementally.
A diagram is said to commute when on selecting any two vertices, any directed path leads to the same result by composition.
Naively checking if all paths in a given graph are path independent could require a number of function equality checks as bad as factorial in the number of nodes, since a path consists of an ordering of nodes. Further, the presence of cycles implies a potentially infinite number of paths. Previous work ~\cite{commutative} has identified an $O(|E|^2|V|^4)$ algorithm to verify that an acyclic diagram commutes.

We present a polynomial time $O(|V|(|E|+|V|)$ algorithm to verify that commutativity of a graph is maintained over the course of online addition, assuming an oracle to check the abstract notion of equality of functions. The number of calls to the equality checking oracle required by the algorithm is asymptotically tight. With an additional $O(|V|^4)$ optimization step, is is possible to use instead the minimal possible number of equality checks, should these checks be expensive. We compare our solution to three baselines: a naive checking of all path pairs, with only special handling for cycles, a check for all path pairs that involve the new edge, and with an algorithm suggested by previous work to solve the batch version of the problem for acyclic diagrams.

@@

First, we formalize the notion of a diagram, drawing inspiration from Murota ~\cite{commutative}.

% A lot of this exposition is from the cited paper, I feel like I need to make this clearer to give proper credit, but not sure how.

We start with a directed graph G=(V,E) with sets attached to vertices and maps, to arcs. Each edge (u, v) in E corresponds to a function that maps elements of u to elements in v.
All the functions form a semigroup F, with multiplication denoted by $\circ$.
A semigroup is an algebraic structure with a set and an associative binary operation. We use it to represent functions in order to capture the one allowed operation- function composition.

@@

Two situation can be imagined- the first, where the existence of the identity function from every node to itself is assumed as would describe most real world situations, and the second, where no such identity needs to exist.

@@

\paragraph{Implementation}
*Copy in python code from implementation? Appendix?*

@@

\subsection{Optimal solution for batch problem}
Murota's main result ~\cite{commutative} is a solution to the question of whether a given acyclic diagram commutes that returns the minimal number of equality checks. The paper describes the following algorithm to find the ($V^2E$ bounded) minimal set of pairs that needs to be checked.

@@

A bilinking is defined to be a parallel pair that is disjoint but for their terminal nodes. The set of all bilinkings is $R_0$.
In an acyclic diagram, if all bilinkings are equal, all parallel pairs must also be equal since they any given pair can be expressed as a composition of bilinkings.

@@

for a set of bilinkings r, the closure function cl is defined as:
cl(r) = $\{ s\in R_0|$ s is linearly dependent on $F(r) \}$.

The closure function on r basically captures all the pairs that can be made by made by composing or "gluing together" the bilinkings in r. 
The aforementioned $\langle\rangle$ function captures the orientation of paths, and can be used to constrain compositions to be in the right direction.

Using these two function we finally define the function $\sigma$ on a set of bilinkings r as
$\sigma(r) = \{s \in R_0 | s\in cl(R\cap \langle s \rangle) \}$.
This is the function used to capture all the pairs whose verification is implied by the verification of pairs in r.

$\sigma$ is used to iteratively check if a given pair is redundant. bilinkings are eliminated until a minimum "spanning" subset is reached.

@@

The solution presented here assumes that there is an implicit identity function on every node, as would be the case in most applications. However, a modified version of the algorithm for diagrams without this assumption is presented in the appendix. The modification is similar in spirit to what was described in the description of the naive baselines.

@@

The try block is executed at most $O(|V|^2)$ times, so that this is the bound on the number of pairs verified.

@@

An efficient implementation of the algorithm can run in $O(|V|(|V|+|E|))$ time at a $O(|V|^3)$ space cost. In such an implementation, path finding from a given source node to all potential sink nodes would be done in a single $O(|V|+|E|)$ breadth first search and the results memoized in $O(|V|^2)$ space.

@@

The run time of the first step is $O(|V|^4)$, and that of the second step is $O(|V|)$, so that the overall bound is $O(|V|^4)$.

@@

Set parallelPairs = new Set();
for S in existingGraph.Nodes():
    for T in existingGraph.Nodes():
        try:
            Path pathWithNewEdge = 
                FindPath(
                    graph: existingGraph, 
                    sourceNode: S,
                    sinkNode: newEdge.Source()) +
                newEdge as Path +
                FindPath(
                    graph: existingGraph, 
                    sourceNode: newEdge.Sink(), 
                    sinkNode: T)
            if (S==T):
                // Cycles are a special case.
                parallelPairs.add(pathWithNewEdge, pathWithNewEdge+pathWithNewEdge)
                for edge in (S.forwardEdges - pathWithNewEdge[0]):
                    parallelPairs.add(edge, edge+pathWithNewEdge)
                for edge in (S.backwardEdges - pathWithNewEdge[-1]):
                    parallelPairs.add(edge, pathWithNewEdge+edge)
            else:
                Path pathInOldGraph = FindPath(
                    graph: existingGraph, 
                    sourceNode: S, 
                    sinkNode: T);
                parallelPairs.add((pathInOldGraph, pathWithNewEdge));
        catch (pathFindingFailed Exception):
            // No comparable pairs from node {S} to node {T} 
            // that need to be checked
            continue
    cycle = findCycleWithTerminalNode(newEdge.Source()):
    if cycle:    
        parallelPairs.add(newEdge, cycle+newEdge)
    cycle = findCycleWithTerminalNode(newEdge.Source()):
    if cycle:    
        parallelPairs.add(newEdge, newEdge+cycle)
output parallelPairs;
\end{verbatim}

The optional minimization algorithm is modified to exclude nodes representing cycles where the source and sink are equal, since the reduction case no longer will apply in the absence of an implicit identity on the terminal node of the cycle.
